---
title: "CCIcompass: From Seurat Object to Ranked Cell-Cell Interactions"
author: "Olha Kholod"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{CCIcompass tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning  = FALSE,
  message  = FALSE
)
```

## Overview

**CCIcompass** computes the **Composite Interaction Score (CIS)** to
prioritise cell-cell interactions (CCIs) from single-cell RNA-seq data.
The CIS aggregates results from six CCI inference methods using
**Rank-Biased Precision (RBP)**, rewarding interactions that are
consistently ranked highly across all methods.

The entire pipeline — LIANA inference, donor filtering, method
intersection, and CIS scoring — runs from a single Seurat object via
`run_pipeline()`.

---

## Installation

```{r install, eval=FALSE}
# 1. Install system dependencies (macOS with Apple Silicon only)
#    Open Terminal and run:
#    xcode-select --install

# 2. Install CRAN packages
install.packages(c("devtools", "BiocManager", "dplyr", "ggplot2",
                   "readr", "tibble", "patchwork", "ggrepel"))

# 3. Install Bioconductor dependencies
BiocManager::install(c("BiocGenerics", "S4Vectors", "SingleCellExperiment",
                       "SummarizedExperiment"))

# 4. Install Seurat (v5)
install.packages("Seurat")

# 5. Install LIANA
devtools::install_github("saezlab/liana")

# 6. Install CCIcompass
devtools::install_github("olhakholod/CCIcompass")
```

---

## Quick start — one function call

```{r quickstart, eval=FALSE}
library(CCIcompass)

# Load your Seurat object
obj <- readRDS("my_seurat_object.rds")

# Run the full pipeline
# Option A: cell-type column already contains donor ID
# e.g. meta.data$celltype.combo == "Epi-Intestine_donor1"
results <- run_pipeline(
  seurat_obj = obj,
  idents_col = "celltype.combo",
  min_donors = 2,
  p          = 0.8
)

# Option B: separate cell-type and donor columns
# e.g. meta.data$celltype == "Epi-Intestine"
#      meta.data$patient_id == "donor1"
results <- run_pipeline(
  seurat_obj = obj,
  idents_col = "celltype",
  donor_col  = "patient_id",
  min_donors = 2
)

# View top results
head(results[, c("source", "target", "ligand", "receptor", "total_rbp_score")])

# Plot
plot_cis_dotplot(results, top_n = 20, title = "Top 20 CCIs")
```

---

## Walkthrough with example data

The package includes a small synthetic dataset (`example_consensus`) that
demonstrates the CIS computation step without needing to run LIANA.

### Load the package and example data

```{r load}
library(CCIcompass)
library(dplyr)
library(ggplot2)

# Load the pre-computed consensus table (output of the intersection step)
data("example_consensus")

glimpse(example_consensus)
```

The `example_consensus` table has one row per
patient × source × target × ligand × receptor combination, with a
per-method score column for each of the six LIANA methods.

### Compute CIS

```{r compute_cis}
cis_results <- compute_cis(example_consensus, p = 0.8)

# Top 10 interactions by CIS
cis_results |>
  select(source, target, ligand, receptor, total_rbp_score) |>
  head(10)
```

The `total_rbp_score` is the CIS. The maximum possible value is **6.0**
(ranked first by all six methods). Interactions from the paper —
MIF-CD74, APP-CD74, HLA-E-KLRC1 — consistently appear at the top.

### Visualise as a dot plot

```{r dotplot, fig.cap="Top 20 CCIs by CIS. Point size = CIS value."}
plot_cis_dotplot(
  cis_results,
  top_n = 20,
  title = "Top 20 cell-cell interactions"
)
```

---

## Understanding the pipeline steps

When you call `run_pipeline()`, six steps execute automatically:

### Step 1 — Run LIANA (six methods)

LIANA runs connectome, logfc, natmi, sca, cellphonedb, and cellchat
on your Seurat object using the CellPhoneDB ligand-receptor resource.

```{r liana-diagram, echo=FALSE, fig.cap="Six methods produce independent ranked lists."}
# Schematic only — no computation
methods <- c("connectome", "logfc", "natmi", "sca", "cellphonedb", "cellchat")
scores  <- c(0.82, 0.71, 0.68, 0.75, 0.79, 0.83)
df_methods <- data.frame(method = factor(methods, levels = methods),
                          score  = scores)
ggplot(df_methods, aes(x = method, y = score, fill = method)) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  ylim(0, 1) +
  labs(title = "Example: MIF-CD74 score per method",
       x = NULL, y = "Raw method score") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### Step 2 — Same-donor filtering

Source and target cells must come from the same donor. This removes
spurious cross-donor signals that can arise when multiple donors are
multiplexed in a single Seurat object.

```{r same-donor, eval=FALSE}
# Cell-type labels must encode the donor ID after an underscore:
# "Epi-Intestine_donor1"  ← source and target must share "donor1"
# Labels from different donors are removed.
```

### Step 3 — Multi-donor filter

Only interactions observed in **2 or more donors** (default) are kept.
This removes interactions driven by a single sample.

### Step 4 — Method intersection

Only interactions identified by **all six** methods are kept. This is
the most stringent filter and ensures high-confidence CCIs advance to
scoring.

### Step 5 — CIS computation

For each method $m$, an interaction at rank $r$ receives:

$$\text{RBP}_m = p^{r-1}$$

where $p = 0.8$ (default). The CIS is the sum across all six methods:

$$\text{CIS} = \sum_{m=1}^{6} p^{r_m - 1}$$

```{r rbp-demo, fig.cap="RBP score decreases with rank. p=0.8 (default)."}
ranks <- 1:20
data.frame(
  rank  = ranks,
  score = 0.8 ^ (ranks - 1)
) |>
  ggplot(aes(x = rank, y = score)) +
  geom_line(colour = "#E5835A", linewidth = 1) +
  geom_point(colour = "#E5835A", size = 2) +
  labs(title = "Rank-Biased Precision (p = 0.8)",
       x = "Rank within method", y = "RBP score") +
  theme_bw(base_size = 10)
```

---

## Parameter guide

| Parameter | Default | When to change |
|-----------|---------|----------------|
| `idents_col` | `"celltype.combo"` | Change to your cell-type column name |
| `donor_col` | `NULL` | Set if donor ID is in a separate column |
| `min_donors` | `2` | Lower to `1` for single-donor datasets |
| `p` | `0.8` | Raise toward 1.0 to weight top ranks more |
| `nperms` | `1000` | Reduce to `100` for quick testing |
| `resource` | `"CellPhoneDB"` | Other options: `"Consensus"`, `"LRdb"` |

---

## Session information

```{r session}
sessionInfo()
```
