---
title: "Computing the Composite Interaction Score (CIS) with CCIcompass"
author: "Olha Kholod"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{CIS Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning  = FALSE,
  message  = FALSE
)
```

# Introduction

**CCIcompass** provides the **Composite Interaction Score (CIS)**, an ensemble
metric for prioritising cell–cell interactions (CCIs) from single-cell RNA-seq
data. The CIS is based on **Rank-Biased Precision (RBP)**, which rewards
interactions that are consistently ranked highly across multiple CCI inference
methods.

This vignette walks through the complete workflow using a small, built-in
example dataset so that you can immediately see expected outputs without
needing your own data.

The full analysis pipeline used in the paper is available in
`scripts/figures/Figure_3_pipeline.R`.

---

# Installation

```{r install, eval=FALSE}
# Install dependencies
install.packages(c("devtools", "dplyr", "ggplot2", "readr", "tibble",
                   "patchwork", "ggrepel"))
devtools::install_github("saezlab/liana")   # required for CCI inference

# Install CCIcompass
devtools::install_github("olhakholod/CCIcompass")
```

---

# Overview of the CIS workflow

```
Seurat object
     │
     ▼
┌─────────────────────────────────────────┐
│  Step 1: Run LIANA (6 methods)          │
│  connectome · logfc · natmi · sca       │
│  cellphonedb · cellchat                 │
└─────────────────────────────────────────┘
     │
     ▼
┌─────────────────────────────────────────┐
│  Step 2: Filter to same-patient pairs   │
│  (source and target from same donor)    │
└─────────────────────────────────────────┘
     │
     ▼
┌─────────────────────────────────────────┐
│  Step 3: Retain CCIs common across      │
│  all donors (≥ N donors per cohort)     │
└─────────────────────────────────────────┘
     │
     ▼
┌─────────────────────────────────────────┐
│  Step 4: Intersect across 6 methods     │
│  (keep only consensus interactions)     │
└─────────────────────────────────────────┘
     │
     ▼
┌─────────────────────────────────────────┐
│  Step 5: Compute CIS (RBP aggregation)  │
│  p = patience parameter (default 0.8)   │
└─────────────────────────────────────────┘
     │
     ▼
┌─────────────────────────────────────────┐
│  Step 6: Final filter (≥ 2 donors)      │
│  and visualisation                      │
└─────────────────────────────────────────┘
```

---

# Load the package and example data

```{r load}
library(CCIcompass)
library(dplyr)
library(ggplot2)

# The built-in example dataset contains pre-computed LIANA outputs
# for a small subset of intestinal epithelial interactions.
# It is a named list with one data frame per method.
data("example_cci_list")

# Inspect the structure
names(example_cci_list)
glimpse(example_cci_list$sca)
```

The example dataset has the structure produced by `liana_wrap()`:

| Column | Description |
|--------|-------------|
| `source` | Sending cell type |
| `target` | Receiving cell type |
| `ligand` | Ligand gene symbol |
| `receptor` | Receptor gene symbol |
| `patient_id` | Donor identifier |
| `*_score` | Method-specific interaction score |

---

# Step 1 – Run LIANA (skip if using example data)

```{r liana, eval=FALSE}
library(liana)
library(Seurat)

# Load your Seurat object
obj <- readRDS("path/to/your_seurat_object.rds")

# Run four rank-based methods
liana_out <- liana_wrap(
  obj,
  idents_col = "celltype.combo",
  method     = c("connectome", "logfc", "natmi", "sca"),
  resource   = "CellPhoneDB"
)

# Run CellPhoneDB (permutation-based)
options(future.globals.maxSize = 10000 * 1024^2)
cpdb_out <- liana_wrap(
  obj,
  idents_col         = "celltype.combo",
  method             = "cellphonedb",
  resource           = "CellPhoneDB",
  permutation.params = list(nperms = 100, parallelize = FALSE, workers = 4),
  expr_prop          = 0.05
)

# Run CellChat
cc_out <- liana_wrap(
  obj,
  idents_col = "celltype.combo",
  method     = "call_cellchat",
  resource   = "CellPhoneDB"
)
```

---

# Step 2 – Filter to same-patient source–target pairs

When multiple donors are multiplexed in one object, cell type labels often
include the donor ID (e.g. `"Epi-Intestine_A26"`). This step ensures we only
keep interactions where source and target cells come from the **same donor**,
avoiding spurious cross-donor signals.

```{r filter_patient}
# Using the example data (patient IDs are embedded in source/target labels)
filtered <- filter_same_patient(example_cci_list$sca)

cat("Rows before filtering:", nrow(example_cci_list$sca), "\n")
cat("Rows after filtering: ", nrow(filtered), "\n")
```

---

# Step 3 – Retain CCIs common across all donors

Keep only interactions that appear in **all** (or a specified number of) donors.
This removes donor-specific noise and retains robust signals.

```{r filter_donors}
donors <- c("A26 (386C)", "A38 (432C)", "A30 (398B)")

common <- filter_common_donors(
  filtered,
  donors  = donors,
  n_req   = length(donors)   # require presence in all 3 donors
)

cat("Unique LR pairs after donor filtering:", n_distinct(common$ligand, common$receptor), "\n")
```

---

# Step 4 – Intersect across all 6 methods

Retain only interactions identified by **all six** inference methods. This
conservative filter ensures only high-confidence CCIs advance to scoring.

```{r intersect, eval=FALSE}
# In practice, repeat Steps 2-3 for each method CSV, then:
common_consensus <- intersect_methods(
  list(
    cellchat    = common_cellchat,
    cellphonedb = common_cellphonedb,
    connectome  = common_connectome,
    logfc       = common_logfc,
    natmi       = common_natmi,
    sca         = common_sca
  )
)
```

For this vignette we use the pre-filtered example data:

```{r load_consensus}
data("example_consensus")   # pre-filtered across 6 methods
glimpse(example_consensus)
```

---

# Step 5 – Compute CIS

The **Composite Interaction Score** is computed as:

$$\text{CIS} = \sum_{m=1}^{M} p^{\,r_m - 1}$$

where $r_m$ is the rank of the interaction within method $m$, and $p$ is the
**patience parameter** (default 0.8) that controls how steeply the score
decreases with rank position. Higher $p$ values give more weight to top-ranked
interactions.

```{r compute_cis}
cis_results <- compute_cis(example_consensus, p = 0.8)

# View top interactions
cis_results %>%
  select(source, target, ligand, receptor, total_rbp_score) %>%
  head(10)
```

---

# Step 6 – Final filter (≥ 2 donors) and visualisation

After computing CIS, apply a final filter to keep only interactions observed in
at least two independent donors. This removes interactions driven by a single
sample.

```{r final_filter}
final <- filter_min_donors(cis_results, min_donors = 2)

cat("Final interactions:", nrow(final), "\n")
```

## Dot plot

```{r dotplot, fig.cap="Top ranked CCIs by CIS. Point size = CIS value; colour indicates tissue of origin."}
plot_cis_dotplot(
  final,
  top_n     = 20,
  x_var     = "ligand_receptor",   # "LR" column (ligand-receptor pair)
  y_var     = "source_target",
  size_var  = "total_rbp_score",
  title     = "Top 20 CCIs – Epithelial Intestine"
)
```

## Venn diagram (cross-tissue)

If you have CIS results from multiple tissues, you can compare shared vs.
tissue-specific interactions:

```{r venn, eval=FALSE}
# Example: compare epithelial CCIs across three tissues
plot_cis_venn(
  list(
    intestine = final_intestine,
    skin      = final_skin,
    uterus    = final_uterus
  ),
  labels     = c("epithelial intestine", "epithelial skin", "epithelial uterus"),
  fill_cols  = c("#E5C4A1", "#C8DFA4", "#A4C8DF")
)
```

## PPI hub-gene network

```{r network, eval=FALSE}
# Requires a STRING interaction TSV for the gene set of interest
plot_ppi_network(
  string_tsv  = "path/to/string_interactions.tsv",
  top_n_label = 10,
  seed        = 42
)
```

---

# Session information

```{r session}
sessionInfo()
```
